*DCVtestkit*: an R package for linearity assessment and analysis of quality control dilution curves

### 2022-10-15 Manuscript/manuscript.qmd
Update the manuscript.

---
title: "*DCVtestkit*: an R package for analysis of quality control dilution curves"
bibliography: bibliography.bib
csl: bioinfo.csl
format: docx
---

@@

In metabolomic and lipidomic workflows, a linear response in dilution series generated from pooled quality control (QC) samples is used to assess the quality of the measurement of molecules measured before further analysis. Most of the currently used software only allow the analyst to repetitively plot, view and analyse the dilution curves one at a time, a tedious and time-consuming process. In addition, summary statistics of dilution curves are limited to the Pearson Correlation Coefficient $R$ and/or Coefficient Of Determination $R^2$, which are insufficient to fully understand the shape of the dilution curves. DCVtestkit aims to provide additional summary statistics for dilution curves, taken from previous publications but which are not implemented in the current software tools. It also helps to reduce the analyst’s workload by analysing many dilution curves automatically, reporting the statistical results in Excel and recording the dilution plots in a pdf file. In addition, it can also create an interactive trellis displayed as a HTML folder for more exploratory analyses.

@@

Dilution curves have been used in many areas of biomedical and pharmaceutical sciences. In the field of metabolomics and lipidomics, dilution curves for each molecule measured, usually in a quality control sample generated by pooling equal aliquots from each study sample, are plotted as proposed by @croixmarie2009 to check if the sample matrix is the same among all samples. Molecules that do not show a linear dilution curve can be rejected before further statistical analysis, in spite of their relevance to the study, as their quantitation in different sample groups might be compromised. A cut off Pearson Correlation Coefficient value of $0.8$ is commonly used to balance the risk of accepting or removing signals related to relevant molecules. As such, many laboratories use commercial software to plot these dilution curves and report their corresponding Pearson Correlation Coefficient value. These include either mass spectrometry vendors’ software or, more in general, programs such as Excel and GraphPad Prism. Nevertheless, these tools have limitations.

While the Pearson Correlation Coefficient $R$ and/or Coefficient Of Determination $R^2$ are popular indicators of how well the dilution points fits a linear line, @sonnergaard2006 estimated that these are not effective standalone numeric parameter to estimate linearity. Using a modified version of Anscombe’s quartet, the paper showed that a high coefficient of determination does not reflect well a curve relationship. Both @croixmarie2009 and @sands2021 support this view, by indicating that an obvious non-linear response at the high and/or low ends of the quality control sample-based dilution curve can still yield a high correlation coefficient. This is in agreement with @sanchez2021 findings of several saturated calibration curves with $R$ values greater than $0.995$.

@@

In addition, with the advancement of the technology used for molecular identification coupled with a high emphasis on high throughput analysis and automation, today’s targeted metabolomics/lipidomics can include more than eight hundred analytes measured by a single method. This number can also increase further in an untargeted approach. In these cases, using general-purpose programs to plot dilution curves for every single molecule may be tedious. Moreover, the reports generated by these programs are static dilution plots coupled with a statistical summary table. Having the analyst to look at five hundred dilution curves individually is time consuming and inefficient for data evaluation.

@@

::: {#fig-DCVtestkit-Workflow layout-ncol=1}

![](images/README-Workflow.png)

@@

To improve the description of the goodness-of-fit for linear dilution curves, a few additional parameters, on top of the popular Pearson Correlation Coefficient $R$, are added in the R package *DCVtestkit*. The first additional parameters are statistical results from Mandel’s Fitting Test (as reported in @andrade2013) in which a low $p$ value gives sufficient evidence that a quadratic model is a better fit than a linear model, indicating the curve may not be linear. Another parameter calculated is the Percent Residual Accuracy (as reported in @logue2018) which ranges from $-\infty$ to $100$. If the curve is linear, the value should be close to $100$. Finally, the software also fits a quadratic model on the dilution curve and calculates the concavity, trying to verify if the dilution curve is predominantly non-linear at high (concavity $<0$) or low (concavity $>0$) concentrations.

Returning to @fig-curves, both the saturated and LOD curves give a much lower Mandel’s Fitting Test $p$ values, denoted as `mandel_p_val`, ($1.66 * 10^{-4}$ and $2.56 * 10^{-3}$ respectively vs $0.38$). Correspondingly, ranging from $-\infty$ to $100$, the Percent Residual Accuracy values, denoted as `pra_linear`, are much lower in the saturated and LOD curves compared to the linear curve ($62.30$ and $74.69$ respectively vs $94.32$).

@@

*DCVtestkit* uses the function `evaluate_linearity` to group the dilution curves. At this moment, two workflows are proposed. Workflow $1$ uses both the Pearson Correlation Coefficient and Percent Residual Accuracy to determine if the dilution curve is linear (labelled as Good Linearity) or not (labelled as Poor Linearity). Workflow $2$ goes one step further by using the Mandel’s Fitting Test and the concavity of the fitted quadratic model to check if the non-linear dilution curve plateaus at low concentrations (labelling them as limit of detection) or high concentrations (labelling them as saturation). Non-linear dilution curves that do not follow these trends are then labelled as Poor Linearity.

::: {#fig-Dilution-Grouping-Workflow layout-ncol=1}

![](images/README-LinearEvaluation.png)

@@

::: {#fig-Simulation-Result layout-ncol=1}

![](images/simulation_results.png)

@@

@fig-Simulation-Result showed that Workflow $2$ is able to identify the saturated and limit of detection curves better than the benchmark workflow. On the other hand, while Workflow $2$ is able to identify less linear curves than the benchmark workflow, it gives a comparable score of `r inline_text(simulation_results, variable = wf2_group, level = "Good Linearity", column = "Linear")`

@@

Nevertheless, while the proposed workflow is not a perfect solution, *DCVtestkit* allows the analyst to change and optimise the threshold values, to adjust the determinants of linearity and saturation of dilution curves according to the user’s needs. More information on how to perform such customisation can be found in the program documentation titled: ["Customise Linearity Evaluation Workflow"](https://slinghub.github.io/DCVtestkit/articles/lineval-customization.html). Furthermore, it is also possible to output these statistical summaries in Excel without the classification results.

@@

::: {#fig-Viewing-Interactive-Curves layout-ncol=1}

![](images/README-TrellisOutput.png)

@@

An example of an interactive viewer created by *DCVtestkit* can be accessed from <https://dcvtestkit-interactive-example.netlify.app/>. Such interactive visualisation was also utilised by @hafen2019 in the exploration of proteomics, metabolomics and lipidomics data in a multi-omics approach. The HTML folder can be compressed as a zip file and shared as a pdf or Excel file. This interactive viewer can also be displayed as a report created from Quarto (https://quarto.org/), an open-source scientific and technical publishing system. The report can be viewed at <https://dcvtestkit-quarto-example.netlify.app/>
R script and result outputs can be found in the Supplementary Material or by accessing this GitHub page <https://github.com/JauntyJJS/DCVtestkit_Output>

@@

A linear response in quality control dilution curves is one of the requirements in metabolomics and lipidomics to assess the suitability of the molecules measured for further analysis. Despite its importance in testing data quality, the shortage of software tools to analyse dilution curves effectively and efficiently prevents researchers from fully analysing data. An R package called *DCVtestkit* is created to rectify this issue by allowing many dilution curves to be analysed quickly and reliably. This is done by the use of alternative statistics, other than the Pearson Correlation Coefficient, to better understand the linearity of the dilution curves, as well as reporting of statistical results and dilution plots that can be distributed to others. It is also able to provide an interactive trellis plot for exploratory data analysis. The R package is available on GitHub <https://github.com/SLINGhub/DCVtestkit> while the documentation and tutorials can be accessed from <https://slinghub.github.io/DCVtestkit/>. As *DCVtestkit* is an opensource software, it is accessible to collaboration works which can help improve its usability and performance in the future.

>>>>

---
title: "*DCVtestkit*: an R package for linearity assessment and analysis of quality control dilution curves"
bibliography: bibliography.bib
csl: bioinfo.csl
format: docx
---

@@

Linearity assessment plays a significant role in the validation of quantitative analytical laboratory procedures. In metabolomic and lipidomic workflows, a linear response in dilution series generated from pooled quality control (QC) samples is used to assess the quality of the measurement of molecules measured before further analysis. Most of the currently used software only allow the analyst to repetitively plot, view and analyse the dilution curves one at a time, a tedious and time-consuming process. In addition, summary statistics of dilution curves are limited to the Pearson Correlation Coefficient $R$ and/or Coefficient Of Determination $R^2$, which are insufficient to fully understand the shape of the dilution curves. DCVtestkit aims to provide additional summary statistics for dilution curves, taken from previous publications but which are not implemented in the current software tools. It also helps to reduce the analyst's workload by analysing many dilution curves automatically, reporting the statistical results in Excel and recording the dilution plots in a pdf file. In addition, it can also create an interactive trellis displayed as a HTML folder for more exploratory analyses.

@@

Linearity assessment is one of the criteria used to evaluate the accuracy and precision of a quantitative measurement procedure in science and engineering. Several applications of linearity assessment can be found in @rodríguez1993 for calibration studies, @needleman1990 for drug analysis and @sicilianidecumis2014 for environmental monitoring sensors.

In the field of metabolomics and lipidomics, dilution curves for each molecule measured, usually in a quality control sample generated by pooling equal aliquots from each study sample, are plotted as proposed by @croixmarie2009 to check if the sample matrix is the same among all samples. Molecules that do not show a linear dilution curve can be rejected before further statistical analysis, in spite of their relevance to the study, as their quantitation in different sample groups might be compromised. A cut off Pearson Correlation Coefficient value of $0.8$ is commonly used to balance the risk of accepting or removing signals related to relevant molecules. As such, many laboratories use commercial software to plot these dilution curves and report their corresponding Pearson Correlation Coefficient value. These include either mass spectrometry vendors' software or, more in general, programs such as Excel and GraphPad Prism. Nevertheless, these tools have limitations.

While the Pearson Correlation Coefficient $R$ and/or Coefficient Of Determination $R^2$ are popular indicators of how well the dilution points fits a linear line, @sonnergaard2006 estimated that these are not effective standalone numeric parameter to estimate linearity. Using a modified version of Anscombe's quartet, the paper showed that a high coefficient of determination does not reflect well a curve relationship. Both @croixmarie2009 and @sands2021 support this view, by indicating that an obvious non-linear response at the high and/or low ends of the quality control sample-based dilution curve can still yield a high correlation coefficient. This is in agreement with @sanchez2021 findings of several saturated calibration curves with $R$ values greater than $0.995$.

@@

In addition, with the advancement of the technology used for molecular identification coupled with a high emphasis on high throughput analysis and automation, today's targeted metabolomics/lipidomics can include more than eight hundred analytes measured by a single method. This number can also increase further in an untargeted approach. In these cases, using general-purpose programs to plot dilution curves for every single molecule may be tedious. Moreover, the reports generated by these programs are static dilution plots coupled with a statistical summary table. Having the analyst to look at five hundred dilution curves individually is time consuming and inefficient for data evaluation.

@@

::: {#fig-DCVtestkit-Workflow layout-ncol="1"}
![](images/README-Workflow.png)

@@

To improve the description of the goodness-of-fit for linear dilution curves, a few additional parameters, on top of the popular Pearson Correlation Coefficient $R$, are added in the R package *DCVtestkit*. The first additional parameters are statistical results from Mandel's Fitting Test (as reported in @andrade2013) in which a low $p$ value gives sufficient evidence that a quadratic model is a better fit than a linear model, indicating the curve may not be linear. Another parameter calculated is the Percent Residual Accuracy (as reported in @logue2018) which ranges from $-\infty$ to $100$. If the curve is linear, the value should be close to $100$. Finally, the software also fits a quadratic model on the dilution curve and calculates the concavity, trying to verify if the dilution curve is predominantly non-linear at high (concavity $<0$) or low (concavity $>0$) concentrations.

Returning to @fig-curves, both the saturated and LOD curves give a much lower Mandel's Fitting Test $p$ values, denoted as `mandel_p_val`, ($1.66 * 10^{-4}$ and $2.56 * 10^{-3}$ respectively vs $0.38$). Correspondingly, ranging from $-\infty$ to $100$, the Percent Residual Accuracy values, denoted as `pra_linear`, are much lower in the saturated and LOD curves compared to the linear curve ($62.30$ and $74.69$ respectively vs $94.32$).

@@

*DCVtestkit* uses the function `evaluate_linearity` to group the dilution curves. At this moment, two workflows are proposed. Workflow $1$ uses both the Pearson Correlation Coefficient and Percent Residual Accuracy to determine if the dilution curve is linear (labelled as Good Linearity) or not (labelled as Poor Linearity). Workflow $2$ goes one step further by using the Mandel's Fitting Test and the concavity of the fitted quadratic model to check if the non-linear dilution curve plateaus at low concentrations (labelling them as limit of detection) or high concentrations (labelling them as saturation). Non-linear dilution curves that do not follow these trends are then labelled as Poor Linearity.

::: {#fig-Dilution-Grouping-Workflow layout-ncol="1"}
![](images/README-LinearEvaluation.png)

@@

::: {#fig-Simulation-Result layout-ncol="1"}
![](images/simulation_results.png)

@@

@fig-Simulation-Result showed that Workflow $2$ is able to identify the saturated and limit of detection curves better than the benchmark workflow. On the other hand, while Workflow $2$ is able to identify less linear curves than the benchmark workflow, it gives a comparable score of `r inline_text(simulation_results, variable = wf2_group, level = "Good Linearity", column = "Linear")`. More details on how the simulated data is created and can be found in this report <https://dcvtestkit-simulation.netlify.app>. Relevant R script to generate this report can be found in the Supplementary Material or by accessing this GitHub page <https://github.com/JauntyJJS/DCVtestkit_Output>

@@

Nevertheless, while the proposed workflow is not a perfect solution, *DCVtestkit* allows the analyst to change and optimise the threshold values, to adjust the determinants of linearity and saturation of dilution curves according to the user's needs. More information on how to perform such customisation can be found in the program documentation titled: ["Customise Linearity Evaluation Workflow"](https://slinghub.github.io/DCVtestkit/articles/lineval-customization.html). Furthermore, it is also possible to output these statistical summaries in Excel without the classification results.

@@

::: {#fig-Viewing-Interactive-Curves layout-ncol="1"}
![](images/README-TrellisOutput.png)

@@

An example of an interactive viewer created by *DCVtestkit* can be accessed from <https://dcvtestkit-interactive-example.netlify.app/>. Such interactive visualisation was also utilised by @hafen2019 in the exploration of proteomics, metabolomics and lipidomics data in a multi-omics approach. The HTML folder can be compressed as a zip file and shared as a pdf or Excel file. This interactive viewer can also be displayed as a report created from Quarto (https://quarto.org/), an open-source scientific and technical publishing system. The report can be viewed at <https://dcvtestkit-quarto-example.netlify.app/>. R script and result outputs can be found in the Supplementary Material or by accessing this GitHub page <https://github.com/JauntyJJS/DCVtestkit_Output>

@@

To verify if a quantitative analytical test method is reliable, it is important to check for linearity. A linear response in quality control dilution curves is one of the requirements in metabolomics and lipidomics to assess the suitability of the molecules measured for further analysis. Despite its importance in testing data quality, the shortage of software tools to analyse dilution curves effectively and efficiently prevents researchers from fully analysing data. An R package called *DCVtestkit* is created to rectify this issue by allowing many dilution curves to be analysed quickly and reliably. This is done by the use of alternative statistics, other than the Pearson Correlation Coefficient, to better understand the linearity of the dilution curves, as well as reporting of statistical results and dilution plots that can be distributed to others. It is also able to provide an interactive trellis plot for exploratory data analysis. The R package is available on GitHub <https://github.com/SLINGhub/DCVtestkit> while the documentation and tutorials can be accessed from <https://slinghub.github.io/DCVtestkit/>. As *DCVtestkit* is an opensource software, it is accessible to collaboration works which can help improve its usability and performance in the future.
